Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-18 paper.tex
Pass on final example

Equation~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g.\ parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. This gives a great deal 
of flexibility, which we illustrate in this section via some examples.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/likratio.png}
    \caption{TODO add A/B and change ``unary regions''.
    Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$.}
    \label{fig:lik-surface}
\end{figure}

Our second example shows a different aspect of this robustness to missing 
data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
obtained when we compute the likelihood of an example simulation,
in which the ARG is at three different levels of completeness
(illustrated in Fig~\ref{fig:lik-surface}B).
We first run a simulation under the SMC
using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
ARG explicitly identifies all of the recombination and common 
ancestor events that occurred in the simulation, and 
we can therefore compute a likelihood under the HKY formulation
(for example, using the \texttt{msprime.log\_arg\_likelihood} 
function). The example full ARG in 
Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
unary''~\citep{wong_general_2023}. Nodes 6 and 7 correspond to the 
left and right parents of the recombination event causing the 
breakpoint at position 536~\cite[we need two nodes to uniquely identify
 the passage of ancestral material through recombination 
events; see][]{baumdicker_efficient_2021}, and therefore present
in only those trees. 
Node 8 is a coalescent node in the right-hand tree, but unary
in the left-hand tree, marking the passage of the lineage through
this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
and compute the likelihood using equation~\eqref{eq:full-lik},
there is very little difference in the likelihood ratio
curve from the exact value with complete input. 
Most importantly, the minimum occurs at the true value of the 
recombination parameter.
When we further simplify
the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
we obtain an ARG in the standard format used by \texttt{msprime}.
Although there is substantial information about the correlation
structure of trees in this representation, we lose important
information about recombination. This is demonstrated in 
Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
is clearly biased.

% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, with a known time to recombination,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the lefthand or
righthand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.

However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.
This approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently. And it further highlights the flexibility
of the algorithm. Simply stated and from a data structure perspective,
we can compute a likelihood under the SMC for any valid tree sequence.
The \texttt{tskit} library provides a wide range
of subsetting operations on tree sequences allowing us to slice
the ARG in time and/or space, as well as to reduce the number of
tracked samples. Any such subset of the tree sequence is itself
a valid tree sequence for which we can again compute a likelihood.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
    slice of the ARG as defined by the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. Number of edges per time slice
    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

>>>>

Eq.~\eqref{eq:full-lik} allows us to compute a likelihood for \emph{any}
collection of nodes (ancestral haplotypes that lived at some time 
$t$ in the past) and edges (a record that child node $c$ 
inherited the genomic interval $[x, y)$ from parent node $p$)
under the SMC. Other than some basic consistency constraints (e.g.\ parents
must be older than children) there are no requirements on the topology
described, and in particular, no requirement that the events from the 
model (the SMC) should be explicitly identified in the graph. Our 
formulation reasons about the events that \emph{could have resulted in}
the observed nodes and edges, and therefore frees us from the requirement
of having to explicitly write those down. This gives a great deal 
of flexibility, which we illustrate in this section via some examples.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/likratio.png}
    \caption{TODO add A/B and change ``unary regions''.
    Log likelihood-ratio curves for ARGs simulated under the SMC with
    various degrees of completeness. The likelihood ratio is computed with
    respect to the true recombination rate. In case of the full ARG the likelihood
    under the SMC has been computed using the exact time to the recombination event.
    Simulation performed with \texttt{msprime}, parameters: $L=1$Mb, $20$ diploid samples,
    $r=2.5e-8$, $N_e=1e4$.}
    \label{fig:lik-surface}
\end{figure}

Our second example shows a different aspect of this robustness to missing 
data. In Fig~\ref{fig:lik-surface}A we can see the likelihood-ratio curves
obtained when we compute the likelihood of an example simulation,
in which the ARG is at three different levels of completeness
(illustrated in Fig~\ref{fig:lik-surface}B).
We first run a simulation under the SMC
using the \texttt{record\_full\_arg} option in \texttt{msprime}. This 
ARG explicitly identifies all of the recombination and common 
ancestor events that occurred in the simulation, and 
we can therefore compute a likelihood under the HKY formulation
(for example, using the \texttt{msprime.log\_arg\_likelihood} 
function). The example full ARG in 
Fig~\ref{fig:lik-surface}B has three nodes that are ``locally
unary''~\citep{wong_general_2023}. Nodes 6 and 7 correspond to the 
left and right parents of the recombination event causing the 
breakpoint at position 536~\cite[we need two nodes to uniquely identify
 the passage of ancestral material through recombination 
events; see][]{baumdicker_efficient_2021}, and therefore present
in only those trees. 
Node 8 is a coalescent node in the right-hand tree, but unary
in the left-hand tree, marking the passage of the lineage through
this ancestral haplotype (similar to node g in Fig~\ref{fig:smc-unary}).
Fig~\ref{fig:lik-surface}A shows that, remarkably, when we 
remove the recombination nodes (e.g., 6/7) but keep the locally unary nodes (e.g., 8)
and compute the likelihood using eq.~\eqref{eq:full-lik},
there is very little difference in the likelihood ratio
curve from the exact value with complete input. 
Most importantly, the minimum occurs at the true value of the 
recombination parameter.
When we further simplify
the ARG to remove all unary nodes~\citep{wong_general_2023,kelleher_efficient_2018} 
we obtain an ARG in the standard format used by \texttt{msprime}.
Although there is substantial information about the correlation
structure of trees in this representation, we lose important
information about recombination. This is demonstrated in 
Fig~\ref{fig:lik-surface}A, where the corresponding likelihood ratio curve
is clearly biased.

\begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{figures/3-slices-whitespace.png}
    \caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant
    for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000,
    (5000:\infty):10 000]$. Remaining parameters: $L=10$kb, $20$ diploid samples,
    $r=2.5e-8$, $148$ edges in ARG. A single $N_e$ value was estimated for each
    slice of the ARG as defined by the true population size change time points.
    The inferred posterior distribution (right) using the SMC-likelihood was
    obtained using a uniform prior. Number of edges per time slice
    (from most recent slice to oldest): $[68, 57, 23]$. Locally unary nodes were
    omitted for clarity.}
    \label{fig:3-arg-slices}
\end{figure}

Our final example concerns [TODO motivate this example; why is 
this interesting/useful?]
% slicing of the ARG
By explicitly associating a single likelihood with each edge,
we can compute the likelihood for any slice of the ARG taken in
time and/or along the genome.
In the case of a fully precise ARG, with a known time to recombination,
this is true as long as the slicing operation does not erase the
information on whether an edge is associated with the lefthand or
righthand segment generated by a recombination event. In the absence
of the exact time to recombination, the integral in
eq.~\eqref{eq:depth} means that the likelihood
for the entire ARG does not factorize into the likelihoods of these
individual slices. More precisely, the likelihood associated with any
edge that is initiated by a recombination breakpoint cannot be factorized
into the product of two integrals of the same form.
However, as long as the individual slices of the ARG contain sufficient
information we can approximate the true integral as the product of the
integral of the slices. Fig.~\ref{fig:3-arg-slices} demonstrates this idea
on a toy example. We simulated an ARG under human-like parameters with
two effective population size changes, and subsequently computed the posterior
distribution on the three effective population sizes
using a uniform prior on each slice independently.
The result is reasonably accurate, despite only using 10kb of sequence.
This approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution and context to a dataset of genetic variation.
It also opens up the opportunity for more efficient optimization
strategies during inference given that each slice of the ARG can be
processed independently.

% And it further highlights the flexibility
% of the algorithm. Simply stated and from a data structure perspective,
% we can compute a likelihood under the SMC for any valid tree sequence.
% The \texttt{tskit} library provides a wide range
% of subsetting operations on tree sequences allowing us to slice
% the ARG in time and/or space, as well as to reduce the number of
% tracked samples. Any such subset of the tree sequence is itself
% a valid tree sequence for which we can again compute a likelihood.

@@
